Indexed Containers

### 2011-05-10 journal/paper.tex
bibs

Using \ensuremath{\Conid{Fin}} and \ensuremath{\Conid{Vec}} instead of \ensuremath{\Conid{Nat}} and \ensuremath{\Conid{List}} enables us to write
a total projection function projecting the nth element out of vector:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Maybe}\;\Conid{A}{}\<[E]%
\\
\>[B]{}[\mskip1.5mu \mskip1.5mu]\;\mathbin{!!}\;\Varid{n}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{nothing}{}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{zero}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{just}\;\Varid{a}{}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{suc}\;\Varid{n}\;{}\<[20]%
\>[20]{}\mathrel{=}\;\Varid{as}\;\mathbin{!!}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Note, that a corresponding function \ensuremath{\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{A}} is not definable in a total language like Agda.

@@

The equality type expresses the focussed character of the
constructors for \ensuremath{\Conid{Fin}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{FNe}\;\in\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{FNe}\;\Conid{X}\;\Conid{Y}\;\Varid{n}\;\mathrel{=}\;\Conid{Fin}\;\Varid{n}\;\uplus\;(\Conid{X}\;\Varid{n}\;\uplus\;\Conid{Y}\;\Varid{n}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{FNf}\;\in\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{FNf}\;\Conid{X}\;\Conid{Y}\;\Varid{n}\;\mathrel{=}\;(\Conid{Y}\;\ensuremath{\mbox{$\circ$}}\;\Varid{suc})\;\Varid{n}\;\uplus\;\Conid{X}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\todo{Discuss mutual case.}

@@

This paper is an expanded and revised version of the LICS paper by the
first and 3rd author \cite{lics}. In the present paper we have
integrated the Agda formalisation in the main development, which in
many instances required extending it. We have made explicit the use of
relative monads which was only hinted at in the conference version
based on the recent work on relative monads \cite{relmon}. We have
also dualized the development to terminal coalgebras which required
the type of paths to be defined inductively instead of recursively as
done in the conference paper (section \ref{sec:termcoalg}).  We
have also formalized the derivation of indexed W-types from ordinary
W-types (section \ref{wifromw}. The derivation of M-types from W-types
(section \ref{sec:mfromw})
was already given in \cite{C-CSPTs} is revisited here exploiting the
indexed W-type derived previously. moreover the development is formalized in
Agda.

@@

With these ingredients we obtain a theory which corresponds to
extensional Type Theory \cite{hofmann-conservativity}.

@@

The category of containers is a full and faithful sub-category of the functor category. We have also shown that the category of containers is cartesian closed (\cite{cie}, and is closed under formation of co-products.

@@

A more serious challenge are mutual inductively (or coinductively)
defined families where one type depends on another
\cite{setzer-forsberg-indind}. A typical example is the syntax of Type
Theory itself which, to simplify, can be encoded by mutually defining
contexts containing terms, types in a given context and terms in a
given type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Con}\;\in\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{Ty}\;\in\;\Conid{Con}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{Tm}\;\in\;(\Gamma\;\in\;\Conid{Con})\;\rightarrow\;\Conid{Ty}\;\Gamma\;\rightarrow\;\Conid{Set}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
In recent work \cite{calco-indind} present a categorical semantics for
this kind of definitions based on dialgebras. However, a presentation
of strictly positive definitions in the spirit of contaienrs is not
yet available.

>>>>

Using \ensuremath{\Conid{Fin}} and \ensuremath{\Conid{Vec}} instead of \ensuremath{\Conid{Nat}} and \ensuremath{\Conid{List}} enables us to write
a total projection function projecting the nth element out of vector:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\{\mskip1.5mu \Varid{n}\;\in\;\Conid{ℕ}\mskip1.5mu\}\;\rightarrow\;\Conid{Vec}\;\Conid{A}\;\Varid{n}\;\rightarrow\;\Conid{Fin}\;\Varid{n}\;\rightarrow\;\Conid{A}{}\<[E]%
\\
\>[B]{}[\mskip1.5mu \mskip1.5mu]\;\mathbin{!!}\;(){}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{zero}\;\mathrel{=}\;\Varid{a}{}\<[E]%
\\
\>[B]{}(\Varid{a}\;\Varid{∷}\;\Varid{as})\;\mathbin{!!}\;\Varid{suc}\;\Varid{n}\;\mathrel{=}\;\Varid{as}\;\mathbin{!!}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Note, that a corresponding function \ensuremath{\Varid{\char95 !!\char95 }\;\in\;\{\mskip1.5mu \Conid{A}\;\in\;\Conid{Set}\mskip1.5mu\}\;\rightarrow\;\Conid{List}\;\Conid{A}\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{A}} is not definable in a total language like Agda.

@@

The equality type expresses the focussed character of the constructors
for \ensuremath{\Conid{Fin}}. The mutual definition of \ensuremath{\Conid{NeLam}} and \ensuremath{\Conid{NfLam}} can be
represented by two binary functors:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{FNe}\;\in\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{FNe}\;\Conid{X}\;\Conid{Y}\;\Varid{n}\;\mathrel{=}\;\Conid{Fin}\;\Varid{n}\;\uplus\;(\Conid{X}\;\Varid{n}\;\times\;\Conid{Y}\;\Varid{n}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{FNf}\;\in\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{FNf}\;\Conid{X}\;\Conid{Y}\;\Varid{n}\;\mathrel{=}\;(\Conid{Y}\;\ensuremath{\mbox{$\circ$}}\;\Varid{suc})\;\Varid{n}\;\uplus\;\Conid{X}\;\Varid{n}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We can construct \ensuremath{\Conid{NeLam}} and \ensuremath{\Conid{NfLam}} by an elimination procedure:
first we define a parametrized initial algebra \ensuremath{\Conid{NeLam'}\;\in\;(\Conid{ℕ}\;\rightarrow\;\Conid{Set})\;\rightarrow\;\Conid{ℕ}\;\rightarrow\;\Conid{Set}} so that \ensuremath{\Conid{NeLam'}\;\Conid{Y}} is the initial algebra of \ensuremath{\lambda\;\Conid{X}\;\rightarrow\;\Conid{FNe}\;\Conid{X}\;\Conid{Y}}
and then \ensuremath{\Conid{NfLam}} is the initial algebra of \ensuremath{\lambda\;\Conid{Y}\;\rightarrow\;\Conid{FNf}\;(\Conid{NeLam'}\;\Conid{Y})\;\Conid{Y}}.
Symmetrically we derive \ensuremath{\Conid{NeLam}} --- compare this with the encoding in
section \ref{sec:spf}.

@@

This paper is an expanded and revised version of the LICS paper by the
first and 3rd author \cite{alti:lics09}. In the present paper we have
integrated the Agda formalisation in the main development, which in
many instances required extending it. We have made explicit the use of
relative monads which was only hinted at in the conference version
based on the recent work on relative monads \cite{alti:fossacs10}. We have
also dualized the development to terminal coalgebras which required
the type of paths to be defined inductively instead of recursively as
done in the conference paper (section \ref{sec:termcoalg}).  We
have also formalized the derivation of indexed W-types from ordinary
W-types (section \ref{wifromw}. The derivation of M-types from W-types
(section \ref{sec:mfromw})
was already given in \cite{alti:cont-tcs} is revisited here exploiting the
indexed W-type derived previously. moreover the development is formalized in
Agda.

@@

With these ingredients we obtain a theory which corresponds to
extensional Type Theory \cite{hofmann1996conservativity}.

@@

The category of containers is a full and faithful sub-category of the functor category. We have also shown that the category of containers is cartesian closed (\cite{txa:cie10}, and is closed under formation of co-products.

@@

A more serious challenge are mutual inductively (or coinductively)
defined families where one type depends on another
\cite{forsberg2010inductive}. A typical example is the syntax of Type
Theory itself which, to simplify, can be encoded by mutually defining
contexts containing terms, types in a given context and terms in a
given type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Con}\;\in\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{Ty}\;\in\;\Conid{Con}\;\rightarrow\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Conid{Tm}\;\in\;(\Gamma\;\in\;\Conid{Con})\;\rightarrow\;\Conid{Ty}\;\Gamma\;\rightarrow\;\Conid{Set}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
In recent work \cite{txa:catind2} present a categorical semantics for
this kind of definitions based on dialgebras. However, a presentation
of strictly positive definitions in the spirit of contaienrs is not
yet available.
